---
title: "Data Science: Inference and Modeling"
subtitle: ""
author: "H. Gjerning"
date: "Jun, 2018"
output:
  pdf_document: default
  html_document: default
data: null
# Source: "https://courses.edx.org/courses/course-v1:HarvardX+PH125.4x+1T2018/progress"
# Destination: "C:\Users\dell2013\Documents\artData\Edx\Harvard\Spec_DataScience\4_InferenceandModeling"
---

### I. Introduction



### II. Data preparation: collection and loading
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, fig_caption = T}

rm(list = ls()) # clears the global environment
setwd("~/artData/Edx/HArvard/Spec_DataScience/4_InferenceandModeling")

```

Next, you will need to install several packages which are required for collecting near real-time tweets and load them into the workspace.
```{r}
# Define the dependency packages we need.
required.packs = c("ggplot2")
        

# Install the required packages if missing, then load them.
sapply(required.packs, function(pack) {
  if(!(pack %in% installed.packages())) {install.packages(pack)}
  require(pack, character.only=TRUE)
})

print("Dependencies met [OK]")

```
####################################################################################
#
#  Module 1: Parameters and Estimates
#
####################################################################################

## 1.1 Parameters and Estimates

Exercise 1. Polling - expected value of S
Suppose you poll a population in which a proportion p of voters are Democrats and 1???p are Republicans. Your sample size is N=25. Consider the random variable S, which is the total number of Democrats in your sample.

What is the expected value of this random variable S?
Answer: E(S)=25p


Exercise 2. Polling - standard error of S
Again, consider the random variable S, which is the total number of Democrats in your sample of 25 voters. The variable p describes the proportion of Democrats in the sample, whereas 1???p describes the proportion of Republicans.

What is the standard error of S?
SE(S)=25p(1???p)???????????????????????????


Exercise 3. Polling - expected value of X-bar
Consider the random variable S/N, which is equivalent to the sample average that we have been denoting as X¯. The variable N represents the sample size and p is the proportion of Democrats in the population.

What is the expected value of X¯?
E(X¯)=p


Exercise 4. Polling - standard error of X-bar
What is the standard error of the sample average, X¯?

The variable N represents the sample size and p is the proportion of Democrats in the population.
SE(X¯)=p(1???p)/N??????????????????????????????


```{r}
# Exercise 5. se versus p

# `N` represents the number of people polled
N <- 25

# Create a variable `p` that contains 100 proportions ranging from 0 to 1 using the `seq` function
p <- seq(0, 1, length = 100)

# Create a variable `se` that contains the standard error of each sample average
se <- sqrt(p*(1-p)/N)

# Plot `p` on the x-axis and `se` on the y-axis
plot(p, se)
```

```{r}
# Exercise 6. Multiple plots of se versus p

# The vector `p` contains 100 proportions of Democrats ranging from 0 to 1 using the `seq` function
p <- seq(0, 1, length = 100)

# The vector `sample_sizes` contains the three sample sizes
sample_sizes <- c(25, 100, 1000)

# Write a for-loop that calculates the standard error `se` for every value of `p` for each of the three samples sizes `N` in the vector `sample_sizes`. Plot the three graphs, using the `ylim` argument to standardize the y-axis across all three plots.
for(N in sample_sizes){
  se <- sqrt(p*(1-p)/N)
  plot(p, se, ylim = c(0,0.5/sqrt(25)))
}
```

Exercise 7. Expected value of d
Our estimate for the difference in proportions of Democrats and Republicans is d=X¯???(1???X¯).

Which derivation correctly uses the rules we learned about sums of random variables and scaled random variables to derive the expected value of d?

E[X¯???(1???X¯)]=E[2X¯???1] =2E[X¯]???1 =2p???1 =p???(1???p)

Exercise 8. Standard error of d
Our estimate for the difference in proportions of Democrats and Republicans is d=X¯???(1???X¯).

Which derivation correctly uses the rules we learned about sums of random variables and scaled random variables to derive the standard error of d?

SE[X¯???(1???X¯)]=SE[2X¯???1] =2SE[X¯] =2p(1???p)/N??????????????????????????????


```{r}
# Exercise 9. Standard error of the spread

# `N` represents the number of people polled
N <- 25

# `p` represents the proportion of Democratic voters
p <- 0.45

# Calculate the standard error of the spread. Print this value to the console.
2*sqrt(p*(1-p)/N)
```

Exercise 10. Sample size
So far we have said that the difference between the proportion of Democratic voters and Republican voters is about 10% and that the standard error of this spread is about 0.2 when N=25. Select the statement that explains why this sample size is sufficient or not.

This sample size is too small because the standard error is larger than the spread.



####################################################################################
#
#  Module 2: The Central Limit Theorem in Practice
#
####################################################################################
In Section 2, you will look at the Central Limit Theorem in practice.

After completing Section 2, you will be able to:

Use the Central Limit Theorem to calculate the probability that a sample estimate  is close to the population proportion .
Run a Monte Carlo simulation to corroborate theoretical results built using probability theory.
Estimate the spread based on estimates of  and .
Understand why bias can mean that larger sample sizes aren't necessarily better.
There is 1 assignment that uses the DataCamp platform for you to practice your coding skills.

We encourage you to use R to interactively test out your answers and further your learning.

## 2.1 Central Limit Theorem in Practice


```{r}


```

## 2.2 Customizing Plots





####################################################################################
#
#  Module 3: Confidence Intervals and p-Values
#
####################################################################################

## 3.1 Confidence Intervals and p-Values


```{r}


```







####################################################################################
#
#  Module 4: Statistical Models
#
####################################################################################

## 4.1 Statistical Models


```{r}


```






####################################################################################
#
#  Module 5: Bayesian Statistics
#
####################################################################################

## 5.1 Bayesian Statistics



```{r}


```


####################################################################################
#
#  Module 6: Election Forecasting
#
####################################################################################

## 6.1 Election Forecasting 




####################################################################################
#
#  Module 7: Association Tests
#
####################################################################################

## 7.1 Association Tests



