---
title: "Data Science: R Basics"
subtitle: ""
author: "H. Gjerning"
date: "Jun, 2018"
output:
  pdf_document: default
  html_document: default
data: null
# Source: "https://www.edx.org/professional-certificate/harvardx-data-science"
# Destination: "C:\Users\dell2013\Documents\artData\Edx\Harvard\Spec_DataScience\1_RBasics"
---

### I. Introduction



### II. Data preparation: collection and loading
```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F, fig_caption = T}

rm(list = ls()) # clears the global environment
setwd("~/artData/Edx/Harvard/Spec_DataScience/1_RBasics")

```

Next, you will need to install several packages which are required for collecting near real-time tweets and load them into the workspace.
```{r}
# Define the dependency packages we need.
required.packs = c("dslabs")
        

# Install the required packages if missing, then load them.
sapply(required.packs, function(pack) {
  if(!(pack %in% installed.packages())) {install.packages(pack)}
  require(pack, character.only=TRUE)
})

print("Dependencies met [OK]")

```
####################################################################################
#
#  Module 1: R Basics, Functions, and Data Types
#
####################################################################################

## 1.1 Motivation

## 1.2 R Basics

```{r}
log(1)
exp(1)


log(exp(1))


```
```{r}
# EX 2: Variable names

# Load package and data

library(dslabs)
data(murders)

# Use the function names to extract the variable names 
names(murders)
```

```{r}
# EX 3: Examining Variables

# To access the population variable from the murders dataset use this code:
p <- murders$population 

# To determine the class of object `p` we use this code:
class(p)

# Use the accessor to extract state abbreviations and assign it to a
a <- murders$abb

# Determine the class of a
class(a)
```

```{r}
# EX 4: Multiple ways to access variables

# We extract the population like this:
p <- murders$population

# This is how we do the same with the square brackets:
o <- murders[["population"]] 

# We can confirm these two are the same
identical(o, p)

# Use square brackets to extract `abb` from `murders` and assign it to b
b <- murders[["abb"]]
# Check if `a` and `b` are identical 
identical(a,b)
```



## 1.3 Data Types


```{r}
class(2)

class("programming")


class(ls)


class(murders)
class(murders$state)
class(murders$region)

# structure
str(murders)

names(murders)

head(murders)
```

```{r}
# EX 2: Variable names

# Load package and data

library(dslabs)
data(murders)

# Use the function names to extract the variable names 
names(murders)
```


```{r}
# EX 5: Factors

# We can see the class of the region variable using class
class(murders$region)

# Determine the number of regions included in this variable 
length(levels(murders$region))
```

```{r}
# EX 6: Tables

# Here is an example of what the table function does
x <- c("a", "a", "b", "b", "b", "c")
table(x)

# Write one line of code to show the number of states per region
table(murders$region)
```


####################################################################################
#
#  Module 2: Vectors, Sorting
#
####################################################################################

## 2.1 Vectors

c stands for concatenate
```{r}
codes <- c(italy=380, canada=124, egypt=818)
codes
```

use [] to access an element of a vector
```{r}
codes[2]

codes[1:2]

codes["canada"]

# codes["egypt","canada"]
```

```{r}
x <- 1:5
x
y <- as.character(x)
y
z <- as.numeric(y)
z
```

```{r}
x <- c("1", "b","3")
x
y <- as.numeric(x)
y
```
```{r}
# EX 1: Numeric Vectors

# Here is an example creating a numeric vector named cost
cost <- c(50, 75, 90, 100, 150)

# Create a numeric vector to store the temperatures listed in the instructions into a vector named temp
# Make sure to follow the same order in the instructions
temp <- c("Beijing", 35, "Lagos", 88, "Paris", 42, "Rio de Janeiro", 84, "San Juan", 81, "Toronto", 30)
temp

temp <- c(35, 88, 42, 84, 81, 30)
```

```{r}
# EX 2: Character vectors

# here is an example of how to create a character vector
food <- c("pizza", "burgers", "salads", "cheese", "pasta")

# Create a character vector called city to store the city names
# Make sure to follow the same order as in the instructions
city <- c("Beijing", "Lagos", "Paris",  "Rio de Janeiro", "San Juan","Toronto")
city
```

```{r}
# EX 3: Connecting Numeric and Character Vectors

# Associate the cost values with its corresponding food item
cost <- c(50, 75, 90, 100, 150)
food <- c("pizza", "burgers", "salads", "cheese", "pasta")
names(cost) <- food

# You already wrote this code
temp <- c(35, 88, 42, 84, 81, 30)
city <- c("Beijing", "Lagos", "Paris", "Rio de Janeiro", "San Juan", "Toronto")

# Associate the temperature values with its corresponding city
names(temp) <- city
temp
```

```{r}
# EX 4: Subsetting vectors

# cost of the last 3 items in our food list:
cost[3:5]

# temperatures of the first three cities in the list:
temp[0:3]
temp[c(1,2,3)]
```

```{r}
# EX 5: Subsetting vectors continued...

# Access the cost of pizza and pasta from our food list 
cost[c(1,5)]

# Define temp
temp <- c(35, 88, 42, 84, 81, 30)
city <- c("Beijing", "Lagos", "Paris", "Rio de Janeiro", "San Juan", "Toronto")
names(temp) <- city

# Access the temperatures of Paris and San Juan
temp[c(3,5)]
```

```{r}
# EX 6: Sequences

# Create a vector m of integers that starts at 32 and ends at 99.
m <- 32:99

# Determine the length of object m.
length(m)

# Create a vector x of integers that starts 12 and ends at 73.
x <- 12:73
# Determine the length of object x.
length(x)
```

```{r}
# EX 7: Sequences continued...

# Create a vector with the multiples of 7, smaller than 50.
seq(7, 49, 7) 

# Create a vector containing all the positive odd numbers smaller than 100.
# The numbers should be in ascending order
seq(1, 99, 2)
```

```{r}
# EX 8: Sequences and length

# We can a vector with the multiples of 7, smaller than 50 like this 
seq(7, 49, 7) 

# But note that the second argument does not need to be last number.
# It simply determines the maximum value permitted.
# so the following line of code produces the same vector as seq(7, 49, 7)
seq(7, 50, 7)

# Create a sequence of numbers from 6 to 55, with 4/7 increments and determine its length
length(seq(6, 55, 4/7))
```

```{r}
# EX 9: Sequences of certain length

# Store the sequence in the object a
a <- seq(1, 10, length.out = 100)

# Determine the class of a
class(a)
```

```{r}
# EX 10: Integers

# Store the sequence in the object a
a <- seq(1, 10)

# Determine the class of a
class(a)
```


```{r}
# EX 11: Integers and Numerics

# Check the class of 1, assigned to the object a
class(1)

# Confirm the class of 1L is integer
class(1L)
```

```{r}
# EX 12: Coercion

# Define the vector x
x <- c(1, 3, 5,"a")

# Note that the x is character vector
class(x)

# Typecast the vector to get an integer vector
# You will get a warning but that is ok
x <- as.integer(x)
```

## 2.2 Sorting
```{r}
# how many murders
sort(murders$total)
```

```{r}
x <- c(31,4,15,92,65)

sort(x)

index <- order(x)
index
```
```{r}
# first we order the total murders and save it to index
index <- order(murders$total)

# then we use index to look up state ordered by murdercount from low to high
murders$abb[index]
```

```{r}
# which is the max murder number
max(murders$total)

# use to look up state
i_max <- which.max(murders$total)
i_max
murders$state[i_max]
```

```{r}
# which is the min murder number
min(murders$total)

# use to look up state
i_min <- which.min(murders$total)
i_min
murders$state[i_min]
```

```{r}
# ranking
rank(x)
```

```{r}
# EX 1: sort

# Access the `state` variable and store it in an object 
states <- murders$state 

# Sort the object alphabetically and redefine the object 
states <- sort(states) 

# Report the first alphabetical value  
states[1]

# Access population values from the dataset and store it in pop
pop <- murders$population
# Sort the object and save it in the same object 
pop <- sort(pop)
# Report the smallest population size 
pop[1]

```

```{r}
# EX 2: order

# Access population from the dataset and store it in pop
pop <- murders$population

# Use the command order, to order pop and store in object o
o <- order(pop)
# Find the index number of the entry with the smallest population size
o[1]


```

```{r}
# EX 3: New Codes

# Find the smallest value for variable total 
which.min(murders$total)

# Find the smallest value for population
which.min(murders$population)
```

```{r}
# EX 4:Using the output of order

# Define the variable i to be the index of the smallest state
i <- which.min(murders$population)

# Define variable states to hold the states
states <- murders$state

# Use the index you just defined to find the state with the smallest population
states[i]
```

```{r}
# EX 5: Ranks
```

```{r}
# EX 5: Ranks

# Store temperatures in an object 
temp <- c(35, 88, 42, 84, 81, 30)

# Store city names in an object 
city <- c("Beijing", "Lagos", "Paris", "Rio de Janeiro", "San Juan", "Toronto")

# Create data frame with city names and temperature 
city_temps <- data.frame(name = city, temperature = temp)

# Define a variable states to be the state names 
states <- murders$state

# Define a variable ranks to determine the population size ranks 
ranks <- rank(murders$population)

# Create a data frame my_df with the state name and its rank
my_df <- data.frame(states = states, ranks = ranks)
my_df
```

```{r}
# EX 6: Data Frames, Ranks and Orders

# Define a variable states to be the state names from the murders data frame
states <- murders$state

# Define a variable ranks to determine the population size ranks 
ranks <- rank(murders$population)

# Define a variable ind to store the indexes needed to order the population values
ind <- order(murders$population)

# Create a data frame my_df with the state name and its rank and ordered from least populous to most 
my_df <- data.frame(states = states[ind], ranks = ranks[ind])
my_df
```

```{r}
# EX 7: NA

# Using new dataset 
library(dslabs)
data(na_example)

# Checking the structure 
str(na_example)

# Find out the mean of the entire dataset 
mean(na_example)

# Use is.na to create a logical index ind that tells which entries are NA
ind <- is.na(na_example)

# Determine how many NA ind has using the sum function
sum(ind)
```
```{r}
# EX 8: Rmoving NAs

# Note what we can do with the ! operator
x <- c(1, 2, 3)
ind <- c(FALSE, TRUE, FALSE)
x[!ind]

# Create the ind vector
library(dslabs)
data(na_example)
ind <- is.na(na_example)

# We saw that this gives an NA
mean(na_example)

# Compute the average, for entries of na_example that are not NA 
mean(na_example[!ind])


```



## 2.3 Vector Arithmetic

```{r}
# which state is the biggest:
murders$state[which.max(murders$population)]

# How many people:
max(murders$population)
```

Example of elementwise operations on vectors
```{r}
# heights in feet
heights <- c(69,62,66,70,70,73,67,73,67,70)
heights * 2.54
```

```{r}
murder_rate <- murders$total/murders$population*100000
murder_rate
```
```{r}
murders$state[order(murder_rate,decreasing=TRUE)]
```

```{r}
# EX 1: Vectorized operations

# Assign city names to `city` 
city <- c("Beijing", "Lagos", "Paris", "Rio de Janeiro", "San Juan", "Toronto")

# Store temperature values in `temp`
temp <- c(35, 88, 42, 84, 81, 30)

# Convert temperature into Celsius and overwrite the original values of 'temp' with these Celsius values

temp <- (temp-32) * 5/9

# Create a data frame `city_temps` 
city_temps <- data.frame(name = city, temperature = temp)
city_temps
```

```{r}
# EX 2: Vectorized operations continued...

# Define an object `x` with the numbers 1 through 100
x <- seq(1, 100)

# Sum the equation 
sum(1/x^2) 
```


```{r}
# EX 3:Vectorized operation continued...

# Load the data
library(dslabs)
data(murders)

# Store the per 100,000 murder rate for each state in murder_rate
murder_rate <- murders$total / murders$population * 100000 
# Calculate the average murder rate in the US 
sum(murder_rate) / length(murder_rate)
mean(murder_rate)
```


####################################################################################
#
#  Module 3: Indexing, Data Wrangeling, Plots
#
####################################################################################

## 3.1 Indexing

```{r}
murder_rate <- murders$total / murders$population * 100000

# murder rate in Italy is 0.71, find us states with similar or lower rates
index <- murder_rate < 0.71
index

# which states
murders$state[index]

# how many states
sum(index)
```

```{r}
# we want to find a states with mountains (West) and safe (murder_rate <= 1)
west <- murders$region == "West"
safe <- murder_rate <= 1
index <- safe & west

murders$state[index]
```

# which
```{r}
x <- c(FALSE, TRUE, FALSE, TRUE, TRUE, FALSE)
which(x)
```
```{r}
# Ex we want to look up the murderrate in Massachusetts
index <- which(murders$state =="Massachusetts")
index

# so to get the murder rate, we use the index
murder_rate[index]
```

```{r}
# Now we want to match severral states
index <- match(c("New York", "Florida", "Texas"), murders$state)
index
# To confirm we got it right
murder_state <- murders$state
murder_state[index]

# and the murder rate of these states
murder_rate[index]
```

```{r}
x <- c("a", "b", "c", "d", "e")
y <- c("a", "d", "f")

# so we can ask if y is in x
y %in% x

```

```{r}
# check if three states are actually states
c("Boston", "Dakota", "Washington") %in% murders$state
```

```{r}
# EX 1: Logical Vectors

# Store the murder rate per 100,000 for each state, in `murder_rate`
murder_rate <- murders$total / murders$population * 100000
# 
# Store the `murder_rate < 1` in `low` 
low <- murder_rate < 1
```

```{r}
# EX 2: which

# Store the murder rate per 100,000 for each state, in murder_rate
murder_rate <- murders$total/murders$population*100000

# Store the murder_rate < 1 in low 
low <- murder_rate < 1

# Get the indices of entries that are below 1
which(low)
```

```{r}
# EX 3: Ordering vectors

# Store the murder rate per 100,000 for each state, in murder_rate
murder_rate <- murders$total/murders$population*100000

# Store the murder_rate < 1 in low 
low <- murder_rate < 1

# Names of states with murder rates lower than 1
murders$state[low]
```

```{r}
# EX 4: Filtering

# Store the murder rate per 100,000 for each state, in `murder_rate`
murder_rate <- murders$total/murders$population*100000

# Store the `murder_rate < 1` in `low` 
low <- murder_rate < 1

# Create a vector ind for states in the Northeast and with murder rates lower than 1. 
ind <- (murders$region == "Northeast") & (murder_rate < 1)

# Names of states in `ind` 
murders$state[ind]
```

```{r}
# EX 5: Filtering continued

# Store the murder rate per 100,000 for each state, in murder_rate
murder_rate <- murders$total/murders$population*100000


# Compute average murder rate and store in avg using `mean` 
avg <- mean(murder_rate)

# How many states have murder rates below avg ? Check using sum 
sum(murder_rate < avg)
```

```{r}
# EX 6: Match

# Store the 3 abbreviations in abbs in a vector (remember that they are character vectors and need quotes)
abbs <- c("AK", "MI", "IA")

# Match the abbs to the murders$abb and store in `ind`
ind <- match(abbs , murders$abb)

# Print state names from `ind`
murders$state[ind]


```

```{r}
# EX 7: %in%

# Store the 5 abbreviations in `abbs`. (remember that they are character vectors)
abbs <- c("MA", "ME", "MI", "MO", "MU")

# Use the %in% command to check if the entries of abbs are abbreviations in the the murders data frame
abbs %in% murders$abb
```

```{r}
# EX 8: Logical operator

# Store the 5 abbreviations in abbs. (remember that they are character vectors)
abbs <- c("MA", "ME", "MI", "MO", "MU") 

# Use the `which` command and `!` operator to find out which abbreviation are not actually part of the dataset and store in ind

ind <- which(!abbs %in% murders$abb)

# What are the entries of abbs that are not actual abbreviations
abbs[ind]
```


## 3.2 Basic Data Wrangeling

```{r}
library(dplyr)
```

we want to add the murder ae into the table
```{r}
murders <- mutate(murders, rate=total/population * 100000)
murders
```

```{r}
filter(murders, rate <= 0.71)
```

select specific columns
```{r}

```


```{r}
new_table <- select(murders,state,region,rate)
new_table
filter(new_table, rate <= 0.71)
```

using pipe to put it all together
```{r}
murders %>% select(state,region,rate) %>% filter(rate <= 0.71)
```

creating frames
```{r}
grades <- data.frame(names=c("John", "juan","Jean","Yao"),
                     exam_1 = c(95, 80, 90, 85),
                     exam_2 = c(90, 85, 85, 90))
grades

class(grades$names)

grades <- data.frame(names=c("John", "juan","Jean","Yao"),
                     exam_1 = c(95, 80, 90, 85),
                     exam_2 = c(90, 85, 85, 90),
                     stringsAsFactors = FALSE)

class(grades$names)
```

```{r}
# EX 1: dplyr

# Loading data
library(dslabs)
data(murders)

# Loading dplyr
library(dplyr)

# Redefine murders so that it includes column named rate with the per 100,000 murder rates
murders <- mutate(murders, rate=total/population * 100000)
```

```{r}
# EX 2: mutate
  
# Note that if you want ranks from highest to lowest you can take the negative and then compute the ranks 
x <- c(88, 100, 83, 92, 94)
rank(-x)

# Defining rate
rate <-  murders$total/ murders$population * 100000

# Redefine murders to include a column named rank
# with the ranks of rate from highest to lowest
murders <- mutate(murders, rank(-rate))
murders
```

```{r}
# EX 3: select

# Load dplyr
library(dplyr)

# Use select to only show state names and abbreviations from murders
select(murders, state, abb)
```

```{r}
# EX 4: filter

# Add the necessary columns
murders <- mutate(murders, rate = total/population * 100000, rank = rank(-rate))

# Filter to show the top 5 states with the highest murder rates
filter(murders, rank <= 5)
```

```{r}
# EX 5: filter with !=

# Use filter to create a new data frame no_south
no_south <- filter(murders, region != "South")

# Use nrow() to calculate the number of rows
nrow(no_south)
```

```{r}
# EX 6: filter with %in%

# Create a new data frame called murders_nw with only the states from the northeast and the west
murders_nw <- filter(murders, region %in% c("Northeast", "West"))

# Number of states (rows) in this category
nrow(murders_nw)

```

```{r}
# EX 7: filtering by two conditions

# add the rate column
murders <- mutate(murders, rate =  total / population * 100000, rank = rank(-rate))

# Create a table, call it `my_states`, that satisfies both the conditions 
my_states <- filter(murders, region %in% c("Northeast", "West") & rate < 1)

# Use select to show only the state name, the murder rate and the rank
select(my_states, state, rate, rank)
```

```{r}
# EX 8: Using the pipe %>%

## Define the rate and rank column
murders <- mutate(murders, rate =  total / population * 100000, rank = rank(-rate))

# show the result and only include the state, rate, and rank columns, all in one line
filter(murders, region %in% c("Northeast", "West") & rate < 1) %>%  
   select(state, rate, rank)

```

```{r}
# EX 9: mutate, filter and select

# Loading the libraries
library(dplyr)
data(murders)

# Create new data frame called my_states (with specifications in the instructions)
my_states <- murders %>% 
    mutate(rate =  total / population * 100000, rank = rank(-rate)) %>%
    filter(region %in% c("Northeast", "West") & rate < 1) %>%
    select(state, rate, rank)

my_states
```


## 3.3 ## Basic Plots

more populus states have more murder
```{r}
population_in_millions <- murders$population
total_gun_murders <- murders$total

plot(population_in_millions, total_gun_murders)
```

To look at the distribution of the data we use histogram
```{r}
class(murders$rate)

murders <- mutate(murders, rate =  total / population * 100000, rank = rank(-rate))
murders

hist(murders$rate)

#One extreme value
murders$state[which.max(murders$rate)]
```

boxplots are good at comparing different groupings like regions
```{r}
boxplot(rate~region, data= murders)
```

```{r}
# EX 1: Scatterplots

# Load the datasets and define some variables
library(dslabs)
data(murders)

population_in_millions <- murders$population/10^6
total_gun_murders <- murders$total

plot(population_in_millions, total_gun_murders)

# Transform population using the log10 transformation and save to object log10_population
log10_population <- log10(murders$population)

# Transform total gun murders using log10 transformation and save to object log10_total_gun_murders
log10_total_gun_murders <- log10(total_gun_murders)

# Create a scatterplot with the log scale transformed population and murders 
plot(log10_population, log10_total_gun_murders)
```

```{r}
# EX 2: Histograms

# Store the population in millions and save to population_in_millions 
population_in_millions <- murders$population/10^6


# Create a histogram of this variable
hist(population_in_millions)
```

```{r}
# EX 3: Boxplots

# Create a boxplot of state populations by region for the murders dataset
boxplot(murders$population~murders$region)


```




####################################################################################
#
#  Module 4: Programming Basics
#
####################################################################################

## 4.1 Introduction to Programming in R






## 4.2 Conditionals

```{r}
library(dslabs)
data(murders)
murder_rate <- murders$total/murders$population*100000

ind <- which.min(murder_rate)

if(murder_rate[ind] < 0.5) {
  print(murders$state[ind])
} else{
  print("No state has murder rate that low")
}

```

```{r}
ind <- which.min(murder_rate)

if(murder_rate[ind] < 0.25) {
  print(murders$state[ind])
} else{
  print("No state has murder rate that low")
}

```

```{r}
a <- c(0,1,2,-4,5)

result <- ifelse(a > 0, 1/a, NA)
result
```

## 4.3 Functions

```{r}
avg <- function(x) {
  s <- sum(x)
  n <- length(x)
  s/n
}

x <- c(5,4,3,2)

avg(x)
```


## 4.4 For Loops

```{r}
compute_s_n <- function(n){
  x <- 1:n
  sum(x)
}

compute_s_n(3) # 1+2+3
compute_s_n(100)

```

we now want to repeat the process 25 times

```{r}
m <- 25
# we create an empty vector
s_n <- vector(length = m)

for(n in 1:m) {
  s_n[n] <- compute_s_n(n)
}

s_n
```

```{r}
n <- 1:m
plot(n, s_n)
```

in stead of loops we use:

  * apply
  * sapply
  * tapply
#

```{r}
# EX 2: Conditionals

# Assign the state abbreviation when the state name is longer than 8 characters 
new_names <- ifelse(nchar(murders$state)>8, murders$abb, murders$state)
new_names
```

```{r}
# EX 4: Defining functions

# Create function called `sum_n`
sum_n <- function(n){
    x <- 1:n
    sum(x)
}

# Determine the sum of integers from 1 to 5000
sum_n(5000)
```

```{r}
# EX 5: Defining functions continued...

# Create `altman_plot` 
altman_plot <- function(x, y){
    plot(x + y, y - x)
}

x <- c(1,2,3,4,5)

y <- c(2,4,6,8,10)

altman_plot(x,y)

```

```{r}
# Run this code 
x <- 3
    my_func <- function(y){
    x <- 5
    y+5
}

# Print value of x 
    
x
```

```{r}
# EX 7: For loops
# Here is a function that adds numbers from 1 to n
example_func <- function(n){
    x <- 1:n
    sum(x)
}

# Here is the sum of the first 100 numbers
example_func(100)

# Write the function with argument n, with the above mentioned specifications and store it in `compute_s_n` 
compute_s_n <- function(n){
  x <- 1:n
  sum(x^2)
}

# Report the value of the sum when n=10
compute_s_n(10)


```

```{r}
# EX 8: For loops continued...

# Define a function and store it in `compute_s_n`
compute_s_n <- function(n){
  x <- 1:n
  sum(x^2)
}

# Create a vector for storing results
s_n <- vector("numeric", 25)

# Assign values to `n` and `s_n`
for(i in 1:25){
  s_n[i] <- compute_s_n(i)
}
```


```{r}
# EX 9: Checking our math

# Define the function
compute_s_n <- function(n){
  x <- 1:n
  sum(x^2)
}

# Define the vector of n
n <- 1:25

# Define the vector to store data
s_n <- vector("numeric", 25)
for(i in n){
  s_n[i] <- compute_s_n(i)
}

#  Create the plot 
plot(n, s_n)
```

```{r}
# EX 10: Checking our math continued

# Define the function
compute_s_n <- function(n){
  x <- 1:n
  sum(x^2)
}

# Define the vector of n
n <- 1:25

# Define the vector to store data
s_n <- vector("numeric", 25)
for(i in n){
  s_n[i] <- compute_s_n(i)
}

# Check that s_n is identical to the formula given in the instructions.
identical(s_n, n*(n+1)*(2*n+1)/6)

```


